---
title: "SporGerm_db"
output: html_document
date: "2025-06-17"
---

```{r}
library(here)
library(tidyverse)
library(UniprotR)
library(data.table)
library(httr)
library(readr)


###print/check working dir of project
here::here()

### call uniprot API functions to submit jobs and print results

source(here("functions/UniprotFunc.R"))

```


### Clean SubtiWiki Gene Data
```{r}
### All genes and functional categories on SubtiWiki
# https://subtiwiki.uni-goettingen.de/v5/data-export/genes

##functional categories
subti.cat <- read.csv(here("data/reference_data/2025-06-17_16-51-14_subti_wiki_export_categories.csv"))

## pull all categories associated with with sporulation/germination (4.2.*) 
### 3.3.7.3 is also associated with sporulation/germination, but it is redundant so we can leave it out
subti.cat <- subset(subti.cat, grepl("^4\\.2\\b", dot_notation))
colnames(subti.cat) <- c("cat_id", "dot_notation", "cat_name")

### label each category as broadly related to sporulation or germination
subti.cat$broad_func <- "sporulation"
subti.cat$broad_func <- ifelse(subti.cat$cat_id %in% c(136, 137, 408, 407), "germination", subti.cat$broad_func)


## load all subtiwiki genes
subti.gene <- read.csv(here("data/reference_data/2025-06-17_16-51-38_subti_wiki_export_genes.csv"))
subti.gene <- subti.gene[,c(3,16,23)]
colnames(subti.gene) <- c("gene_name", "gene_locus", "gene_cat")


library(stringr)

### clean subtiwiki formating 

### remove everything but the subtiwki categories, which are inside the paranthesis 
extr<- str_extract_all(subti.gene$gene_cat, "\\(\\d+\\)")

#get rid of paranthesis and put the clean string back into subti.gene as a new column (subticats), that lists all gene categories assigned to each gene
subti.gene$SubtiCats <- sapply(extr, function(x) paste(str_remove_all(x, "[()]"), collapse = ", "))

## create a row for each cattegory
subti.gene$cat_id <- subti.gene$SubtiCats
subti.gene <- separate_rows(subti.gene, cat_id, sep = ",\\s*")


### merge all sporulation/germ cats with SubtiWiki genes and keep only hits

subti <- merge(subti.cat, subti.gene, by.x="cat_id", by.y="cat_id", all.x=FALSE, all.y=FALSE)


## note that some genes have multiple spor/germ categories, so there are duplicate "genes" in here.
## remove "duplicates" by dropping the specific categories and grouping by either spor / germ
subti.clean <- subti[,c(4,5,6,8)]

subti.clean <- unique(subti.clean)

## any remaining duplicate genes are genes that are associated with both germ and spor
both_germspor <- as.data.frame(subti.clean$gene_locus[duplicated(subti.clean$gene_locus)])
colnames(both_germspor) <- "gene_locus"
both_germspor$subti_func <- "sporulation_germination"

### merge in the sporulation_germination category
subti.clean <- merge(subti.clean, both_germspor, by="gene_locus", all.x=TRUE, all.y=FALSE)
subti.clean$subti_func[is.na(subti.clean$subti_func)] <- "NA"
subti.clean$subti_func <- ifelse(subti.clean$subti_func=="NA", subti.clean$broad_func, subti.clean$subti_func)

### clean subti 
subti.clean <- subti.clean[,-c(2)]
subti.clean <- unique(subti.clean)

```


### prepare subtiwiki genes for submission to Uniprot
```{r}

## Create alternate versions of gene loci without underscores so both versions can be searched. Only  one version should hit to a result. 

subti.clean$gene_locus2 <- gsub("_", "", subti.clean$gene_locus)

## Create a single list of loci name 

df1 <- as.data.frame(subti.clean$gene_locus)
df2 <- as.data.frame(subti.clean$gene_locus2)
colnames(df1) <- "gene_locus"
colnames(df2) <- "gene_locus"

bsub.loci2 <- rbind(df1, df2)

bsub.loci <- unique(bsub.loci2[,c(1)])


#### write list of B. sub gene names associated with sporulation/germination. 

write.table(bsub.loci, here("data/output_data/bsub_loci_sporgerm.txt"), row.names=FALSE, col.names=FALSE, quote=FALSE)

```

### map bsub.loci to Uniprot

```{r}
## Using above list as input, map these gene loci to their UniProt IDs using https://www.uniprot.org/id-mapping and their supplied API code (functions in R source). 
## This can also be done on their website with the settings: Gene_Name → UniProtKB


### load IDs
files = list(
  ids = paste(bsub.loci, collapse = ","),
  from = "Gene_Name",
  to = "UniProtKB"
)



r <- POST(url = "https://rest.uniprot.org/idmapping/run", body = files, encode = "multipart", accept_json())
submission <- content(r, as = "parsed")

### url_fields: list columns/metadata you want from uniprot
## see: https://www.uniprot.org/help/return_fields 
## you will be re-requesting this data with additional metadata later in the code, so no need to add additional url_fields requirements in this code chunk

url_fields <- c("accession", "reviewed", "protein_name","gene_names")
url_fields <- paste(url_fields, collapse = ",")


if (isJobReady(submission[["jobId"]])) {
  url <- paste("https://rest.uniprot.org/idmapping/details/", submission[["jobId"]], sep = "")
  r <- GET(url = url, accept_json())
  details <- content(r, as = "parsed")
  url <- getResultsURL(details[["redirectURL"]])
  # Using TSV format see: https://www.uniprot.org/help/api_queries#what-formats-are-available
  url <- paste(url, "?fields=accession,gene_names,keyword&format=tsv", sep = "")
  print(url)
  r <- GET(url = url, accept_json())
  bsub.prot = read.table(text = content(r), sep = "\t", header=TRUE)
 
}


### high amount of discrepancy between number of rows in  bsub.loci and bsub.prot is expected, since bsub.loci includes alternate versions of the same gene name because some gene names on Uniprot are formmated with underscores while others are not. So the easiest to make sure all are captured by the search is to include both potential versions in bsub.loci (BSUB_XXXX and BSUBXXXX) -- but only one of these will actually hit for each protein in bsub.prot.
```


```{r}

## return Subti gene loci to original Subti format to remerge with Subti metadata
bsub.prot$From <- ifelse(!grepl("_", bsub.prot$From), gsub("BSU","BSU_",bsub.prot$From), bsub.prot$From)

bsub <- merge(subti.clean, bsub.prot, by.y="From", by.x="gene_locus", all=TRUE )

subtiWiki.Uni.key <- bsub[,c(6,1:4)]

##pull Uniprot entry ID for use later
bsub.ID <- as.data.frame(bsub$Entry)

```


### Find all Sporulation Hits in Uniprot
```{r}


# Define URL of query, this only returns IDs and entry names of swissprot reviewed eubacteria domain, tagged with sporulation: KW-0749 or germination: KW-0309 keyword

## URL of search query in Uniprot: https://www.uniprot.org/uniprotkb?query=%28taxonomy_id%3A2%29+AND+%28reviewed%3Atrue%29+AND+%28keyword%3AKW-0749%29+OR+%28taxonomy_id%3A2%29+AND+%28reviewed%3Atrue%29+AND+%28keyword%3AKW-0309%29


## modified url of search query to only return gene_name and accession number as metadata

base_url <- "https://rest.uniprot.org/uniprotkb/stream?compressed=true&fields=accession%2Cid%2Cgene_names&format=tsv&query=%28taxonomy_id%3A2%29+AND+%28reviewed%3Atrue%29+AND+%28keyword%3AKW-0749%29+OR+%28taxonomy_id%3A2%29+AND+%28reviewed%3Atrue%29+AND+%28keyword%3AKW-0309%29"


# send GET request to Uniprot
response <- GET(url = base_url, accept("application/gzip"))

# check response for completion, then turn response into data
if (status_code(response) == 200) {
  # Read compressed response as raw binary
  raw_data <- content(response, "raw")
  
  # Use gzcon() to create a readable text connection
  con <- gzcon(rawConnection(raw_data))
  
  # Read TSV data
  KWSpor <- read_tsv(con, col_types = cols(.default = "c"))  
  close(con)  # Close connection
  
  # Print first few rows
  print(head(KWSpor))
  
  # Save as a TSV file
  write_tsv(KWSpor, here("data/output_data/id_uniprot_results.tsv"))
  
} else {
  print(paste("Error:", status_code(response)))
}



keySpor.ID <- as.data.frame(KWSpor$Entry)
```

#### at some point, will need to rewrite code / add a second .Rmd to add in any manual additions from Daniel 2023, other hand-curated proteins, or any subtiWiki proteins which didn't hit using the BSU_locus method.

```{r}


### create vector of all Uniprot and BsubIDs


colnames(keySpor.ID) <-"Entry"
colnames(bsub.ID) <-"Entry"

Uni.ID <- rbind(keySpor.ID, bsub.ID)


Uni.ID <- unique(Uni.ID) ###1290 unique IDs flagged as sporulation either by Subtiwiki or Uniprot Keyword
Uni.ID <- subset(Uni.ID, Uni.ID!=is.na(Uni.ID))
vec.ID <- Uni.ID$Entry


```



### GET ALL IDS in one file with metadata

```{r}
### load IDs
files = list(
  ids = paste(vec.ID, collapse = ","),
  from = "UniProtKB_AC-ID",
  to = "UniProtKB")


r <- POST(url = "https://rest.uniprot.org/idmapping/run", body = files, encode = "multipart", accept_json())
submission <- content(r, as = "parsed")

### url_fields: list columns/metadata you want from uniprot 
## see: https://www.uniprot.org/help/return_fields 
## xref_ko is no longer a valid entry on uniprot

url_fields <- c("accession", "reviewed", "protein_name","gene_names", "annotation_score","keyword","lit_pubmed_id", "gene_orf", "xref_kegg", "xref_eggnog", "xref_refseq", "xref_cazy", "xref_brenda", "xref_orthodb", "xref_pfam","xref_interpro", "xref_biocyc", "xref_unipathway")
url_fields <- paste(url_fields, collapse = ",")


if (isJobReady(submission[["jobId"]])) {
  url <- paste("https://rest.uniprot.org/idmapping/details/", submission[["jobId"]], sep = "")
  r <- GET(url = url, accept_json())
  details <- content(r, as = "parsed")
  url <- getResultsURL(details[["redirectURL"]])
  # request metadata fields + tsv format by adding it to end of job url
  url <- paste(url, "?fields=", url_fields, "&format=tsv", sep = "")
  print(url)
  r <- GET(url = url, accept_json())
  raw_text <- content(r, as = "text")
  Master.List <- fread(input = raw_text, sep = "\t", header = TRUE, quote = "")
  Master.List <- as.data.frame(Master.List)
}



### add in the SubtiWiki metadata fro relevant proteins
Master.Metadata <- merge(Master.List, subtiWiki.Uni.key, by="Entry", all.x=TRUE, all.y=FALSE)



```


```{r}

### search UniProt Keywords for sporulation, germination, or both and assign respective category to UniKey column.

Master.Metadata$UniKey <- "NA"

Master.Metadata$UniKey <- ifelse(grepl("sporulation", Master.Metadata$Keywords, ignore.case = TRUE), "sporulation", Master.Metadata$UniKey)


Master.Metadata$UniKey <- ifelse(grepl("germination", Master.Metadata$Keywords, ignore.case = TRUE), "germination", Master.Metadata$UniKey)

Master.Metadata$UniKey <- ifelse(grepl("germination", Master.Metadata$Keywords, ignore.case = TRUE) & grepl("sporulation", Master.Metadata$Keywords, ignore.case = TRUE), "sporulation_germination", Master.Metadata$UniKey)



write.csv(Master.Metadata, here("data/output_data/UniprotSporGerm_Master.csv"))

```


### At this point, Master.List is a complete list of all Uniprot IDs flagged by Uniprot or SubtiWiki
### Hand curated  Schwartz et al. 2023 as relating to sporulation still needs to be added in.


### download protein (amino acid) fasta files

```{r}
files = list(
  ids = paste(vec.ID, collapse = ","),
  from = "UniProtKB_AC-ID",
  to = "UniProtKB")


r <- POST(url = "https://rest.uniprot.org/idmapping/run", body = files, encode = "multipart", accept_json())
submission <- content(r, as = "parsed")


if (isJobReady(submission[["jobId"]])) {
  url <- paste("https://rest.uniprot.org/idmapping/details/", submission[["jobId"]], sep = "")
  r <- GET(url = url, accept_json())
  details <- content(r, as = "parsed")
  url <- getResultsURL(details[["redirectURL"]])
  url <- paste(url, "?compressed=false&format=fasta", sep = "")  # Ensure compression is OFF

  print(url)
  
  r <- GET(url = url, accept_json())
  raw_text <- content(r, as = "text")  # Get raw FASTA text

  fasta.out <- strsplit(raw_text, "\n")[[1]]  # Split into lines
  
  ### Clean headers
  fasta.out <- gsub(">sp\\|", ">", fasta.out)
  fasta.out <- gsub("\\|", " ", fasta.out)

  ### Reformat FASTA: one header, one sequence per entry
  formatted_fasta <- c()
  current_seq <- ""
  
  for (line in fasta.out) {
    if (startsWith(line, ">")) {
      # Store previous sequence if it exists
      if (current_seq != "") {
        formatted_fasta <- c(formatted_fasta, current_seq)
      }
      formatted_fasta <- c(formatted_fasta, line)  # Add new header
      current_seq <- ""  # Reset sequence
    } else {
      current_seq <- paste0(current_seq, line)  # Concatenate sequence
    }
  }
  
  # Add last sequence
  if (current_seq != "") {
    formatted_fasta <- c(formatted_fasta, current_seq)
  }

  ### Write to file
  writeLines(formatted_fasta, here("data/output_data/uniprot_sporgerm.faa"))
}

```


